formula1

### 2020-04-01 inst/formula1.Rmd
Hamilton in the wet

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

@@

Above, we see a small part of the Placket-Luce likelihood function for
the 2017 results.  We can find the maximum likelihood estimator by
using the `maxp()` function:

```{r,maxp2017}
mL2017 <- maxp(L2017)
mL2017
dotchart(mL2017,pch=16,main='2017 Formula 1')
dotchart(log(mL2017)[1:23],pch=16,main='2017 Formula 1, log scale')
```

@@

m <- mL2017   # maxp(L2017)
ox <- rank(-points) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

@@

## Null of equal strengths:

```{r,equalstrength}
L1 <- loglik(indep(equalp(L2017)),L2017)
L2 <- loglik(indep(mL2017),L2017)   # mL2017 == maxp(L2017)
pchisq(2*(L2-L1),df=25-1,lower.tail=FALSE)
```

@@

We can consider Hamilton and ask various questions about his playing
strength.  One natural null would be that Hamilton's strength is equal
to the mean strength, or $\frac{1}{25}$.

* $H_2\colon p_\mathrm{Hamilton}\geq\frac{1}{25}$
* $H_3\colon p_\mathrm{Hamilton} < \frac{1}{25}$

Noting that the evaluate for Hamilton's strength
$\widehat{p_\mathrm{Hamilton}}$ is about 0.22, we may consider the
likelihood for $H_2$ to be equal to the unconstrained maximum
likelhood as in `L2` above.  The likelihood for $H_3$ requires a
constrained optimization:

```{r,constham}
small <- 0.01
L3 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,rep(0,23))),fcv=-1/25,give=TRUE)
L3
```

(note that the constrained maximum attains the constraint of
$p_\mathrm{Hamilton}=\frac{1}{25}$).   The likelihood ratio test gives

```{r,liktestivediff}
L2 - L3$value 
```

or a p-value of

```{r,liktestive}
pchisq(2*(L2 - L3$value),df=1,lower.tail=FALSE)
```

@@

```{r,L4like}
small <- 0.01
L4 <- maxp(L2017, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
L4
```

Calculating the support and asymptotic p-values:
```{r,l4val}
L2 - L4$value 
pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
```

@@

```{r,fouryearsdata}
F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ordertable2supp
a <- list(F1_2014, F1_2015, F1_2016, F1_2017)
alldrivers <- all_pnames(a)

@@

```{r,findmallyears}
mallyears <- maxp(F1_total)
```

@@

Taking 2017 as an example:

`
venue pole conditions\
AUS Hamilton   dry \
CHN Hamilton   wet \
BHR Bottas     dry \
RUS Vettel     dry \
ESP Hamilton   dry \
MON Raikkonen  dry \
CAN Hamilton   dry \
AZE Hamilton   dry \
AUT Bottas     dry \
GBR Hamilton   dry \
HUN Vettel     dry \
BEL Hamilton   dry \
ITA Hamilon    dry \
SIN Vettel     wet \
MAL Hamilton   dry \
JPN Hamilton   dry \
USA Hamilton   dry \
MEX Vettel     dry \
BRA Bottas     dry \
ABU Bottas     dry \
`

>>>>

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

@@

Above, we see a small part of the Placket-Luce likelihood function for
the 2017 results.  But from now on we will use built-in dataset
`formula1`, which contains extra information including non-finishing
competitors.  We can find the maximum likelihood estimator by using
the `maxp()` function:

```{r,maxp2017,cache=TRUE}
rm(L2017)
mL2017 <- maxp(formula1)
mL2017
dotchart(mL2017,pch=16,main='2017 Formula 1')
dotchart(log(mL2017)[1:23],pch=16,main='2017 Formula 1, log scale')
```

@@

m <- mL2017   # maxp(formula1)
ox <- rank(-points) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

@@

## Hypothesis: all players have equal strengths

Here we write $H_0\colon p_1=p_2=\cdots =p_n=\frac{1}{n}$.

```{r,equalstrength,cache=TRUE}
equalp.test(formula1)
```

@@

We can consider Hamilton and ask various questions about his playing
strength.  One natural hypothesis would be that Hamilton's strength is
equal to the mean strength, or $\frac{1}{25}$.

* $H_A\colon\sum p_i=1$
* $H_0\colon p_\mathrm{Hamilton} < \frac{1}{25}$

(we attempt to reject the null $H_0$).  Noting that the evaluate for
Hamilton's strength $\widehat{p_\mathrm{Hamilton}}$ is about 0.29, we
ask how much support is lost, relative to the support at the
unconstrained evaluate.

```{r,specifichamiltonstrength,cache=TRUE}
specificp.test(formula1,"Hamilton")  # default value is 1/25=0.04
```

(note that the constrained maximum is on the boundary of the
admissible region with $p_\mathrm{Hamilton}\approxequal\frac{1}{25}$).

@@

```{r,L4like,cache=TRUE}
small <- 0.01
L4 <- maxp(formula1, startp=c(small, rep((1-small)/24,23)),fcm=rbind(c(-1,1,rep(0,22))),fcv=0,give=TRUE)
L4
```

Calculating the support and asymptotic p-values:
```{r,l4val}
#L2 - L4$value 
#pchisq(2*(L2 - L4$value),df=1,lower.tail=FALSE)
2+3
```

@@

```{r,fouryearsdata, cache=TRUE}
F1_2014 <- "formula1_2014.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2015 <- "formula1_2015.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2016 <- "formula1_2016.txt" %>% read.table(header=TRUE) %>% ordertable2supp
F1_2017 <- "formula1_2017.txt" %>% read.table(header=TRUE) %>% ordertable2supp
a <- list(F1_2014, F1_2015, F1_2016, F1_2017)
alldrivers <- all_pnames(a)

@@

```{r,findmallyears,cache=TRUE}
mallyears <- maxp(F1_total)
```

@@

```{r,loadtable}
(details <- read.table("formula1_conditions.txt",header=TRUE))
drivers <- pnames(formula1)
```

We cannot use `ordertable2supp()` here because we somehow need to
incorporate pole position and conditions into our likelihood function.
The first step is to modify `f2017` so that non-finishers have a zero
position:

At this point, we introduce a new player: `Ham_wet`, for
"Hamilton in the wet", and we can test the hypothesis that the
`Hamilton == Ham_wet`.  We will cycle through all the venues and,
if it is wet (according to dataframe `conditions`) we will credit
Hamilton's placing to `Ham_wet`.

```{r,hamiltonwetanddry,cache=TRUE}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
f2017 <- f2017[,-ncol(f2017)]
details <- read.table("formula1_conditions.txt",header=TRUE)

drivers <- rownames(f2017)
alldrivers <- c("Ham_dry","Ham_wet",drivers[-1])

L2017 <- hyper2(pnames=alldrivers) # uninformative
for(i in seq_len(ncol(f2017))){
	venue <- colnames(f2017)[i]
	conditions <- details[which(details$venue == venue),4]
	o <- f2017[,i]
        suppressWarnings(o <- as.numeric(levels(o))[o])
        o[is.na(o)] <- 0
	names(o) <- rownames(f2017)
	if(conditions == "wet"){
	   names(o)[names(o) == "Hamilton"] <- "Ham_wet"
	} else if(conditions == "dry"){
	   names(o)[names(o) == "Hamilton"] <- "Ham_dry"
	} else {
	   stop("neither wet nor dry?")
	}
	L2017 <- order_obs(L2017,o)
}  # for(i) loop closes
head(L2017)
```

We can now test the hypothesis that Hamilton's performance in the wet
is the same as his performance in the dry:

```{r,hamwetdrytest,cache=TRUE}
options(width = 80)
samep.test(L2017,c("Ham_wet","Ham_dry"))
```
